Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2014-04-16 acm/ieee_paper_final.tex
minor

\begin{abstract}
In this paper, we develop an efficient scheme for the calculation of
derivatives within the context of Radial Basis Function
Finite-Difference (RBFFD). RBF methods express functions as a linear
combination of \ge{spherically} symmetric basis functions on an arbitrary set of
nodes. The Finite-Difference component expresses this combination over
a local set of nodes neighboring the point where the derivative is
sought.  The derivative at all points takes the form of a sparse
matrix/vector multiplication (SpMV).

@@

\ee{The modelization of Section~\ref{sec:model} establishes that the 
  memory
  transfer caps the computation at 210~GFlop/s, which is 10\% of the
  peak single precision of the architecture. To reach this value, the
  cores need to execute one "useful" Fused Multiply-Add operation
  every ten cycles. The inner loop of the kernel spends 7 vector
  instructions to layout the data and then 3 per tensor product for a
  total of 12 vector instructions. Most of these instruction have a
  one cycle throughput cost except the gather operations which might
  need to perform up to 16 memory accesses. There are also some scalar
  operations but there are few of them and are executed on a different
  pipeline than the vector operations. Overall, the inner loop can
  emit the 4 Fused Multiply-Add in less than 40 cycles.}

@@

  \caption{Different sparsity distributions. In all cases, there are
    512 rows and 32 nonzeros per row. The last two matrices
    corresponds to a derivative stencil in 3D RBFFD
    calculations (with and without RCM ordering).}
  \label{fig:spy_plots}
\end{figure*}

\begin{figure*}
% Figure 9
  \centering
% was width=.48
  \subfigure[Supercompact and Random.]{\includegraphics[width=.35\linewidth]{figures/random_supercompact.pdf}\label{fig:perf_ideal}}
% was width=.48
  \subfigure[Compact and RBFFD.]{\includegraphics[width=.35\linewidth]{figures/rbf_compact.pdf}\label{fig:perf_realistic}}

@@

The different matrices used to perform our experiments and analysis
are presented in Figure~\ref{fig:spy_plots}. The
\ttt{Supercompact} matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. As a result, only 32 values of {\tt x} are read
from memory, and the expense associated with cache misses is removed from consideration. The cost associated with 
{\tt A} remains. The \ttt{Compact} matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for many
applications that rely on sparse matrix vector multiplication. The nonzero
elements of the \ttt{Random}
matrices (Figure~\ref{fig:random}) see their nonzero elements
randomly (uniformly) distributed in the matrix; they represent the
worst case scenario for a cache-based architecture where the cache
reutilisation is the lowest from one row to the next.
The other type of matrix is used in derivative stencils in 3D
RBFFD calculations (Figure~\ref{fig:rbf3dnorcm} shows a $32$-point
stencil of a 3D $8^3$ grid).

@@

In the previous sections, we have explored the practical
implementation on an Intel Xeon Phi card of multiple derivative
operators acting on multiple vectors within the context of
RBFFD. Each derivative has an associated sparse matrix with a fixed
number of nonzeros on each row. While computing a single derivative of
multiple functions is rather common, we accelerate the algorithm further 
by considering multiple
matrices, with identical sparsity patterns, acting on multiple vectors. 
We specialize our study to four matrices and four derivatives,
with 16 outputs, computed as a sum of outer products.

@@

In future work, we will examine the effect of larger stencil sizes and
double precision. The presented performance does not reach the
predicted peak and we believe that common techniques from SpMV such as
cache partitioning can be successfully applied. We will also integrate
the techniques presented in this paper within a fluid simulation using
RBFFD. \ge{We will also look into developing a similar technique on the Kepler
family of GPUs, which offers swizzling and channeling operations.}

>>>>

\begin{abstract}
In this paper, we develop an efficient scheme for the calculation of
derivatives within the context of Radial Basis Function
Finite-Difference (RBF-FD). RBF methods express functions as a linear
combination of \ge{spherically} symmetric basis functions on an arbitrary set of
nodes. The Finite-Difference component expresses this combination over
a local set of nodes neighboring the point where the derivative is
sought.  The derivative at all points takes the form of a sparse
matrix/vector multiplication (SpMV).

@@

\ee{The model of Section~\ref{sec:model} establishes that the 
  memory
  transfer caps the computation at 210~GFlop/s, which is 10\% of the
  peak single precision of the architecture. To reach this value, the
  cores need to execute one "useful" Fused Multiply-Add operation
  every ten cycles. The inner loop of the kernel spends 7 vector
  instructions to layout the data and then 3 per tensor product for a
  total of 12 vector instructions. Most of these instruction have a
  one cycle throughput cost except the gather operations which might
  need to perform up to 16 memory accesses. There are also some scalar
  operations but there are few of them and are executed on a different
  pipeline than the vector operations. Overall, the inner loop can
  emit the 4 Fused Multiply-Add in less than 40 cycles.}

@@

  \caption{Different sparsity distributions. In all cases, there are
    512 rows and 32 nonzeros per row. The last two matrices
    corresponds to a derivative stencil in 3D RBF-FD
    calculations (with and without RCM ordering).}
  \label{fig:spy_plots}
\end{figure*}

\begin{figure*}
% Figure 9
  \centering
% was width=.48
  \subfigure[Supercompact and Random.]{\includegraphics[width=.35\linewidth]{figures/random_supercompact.pdf}\label{fig:perf_ideal}}
% was width=.48
  \subfigure[Compact and RBF-FD.]{\includegraphics[width=.35\linewidth]{figures/rbf_compact.pdf}\label{fig:perf_realistic}}

@@

The different matrices used to perform our experiments and analysis
are presented in Figure~\ref{fig:spy_plots}. The
\ttt{Supercompact} matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. As a result, only 32 values of {\tt x} are read
from memory, and the expense associated with cache misses is removed from consideration. The cost associated with 
{\tt A} remains. The \ttt{Compact} matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for many
applications that rely on sparse matrix vector multiplication. The nonzero
elements of the \ttt{Random}
matrices (Figure~\ref{fig:random}) see their nonzero elements
randomly (uniformly) distributed in the matrix; they represent the
worst case scenario for a cache-based architecture where the cache
reutilisation is the lowest from one row to the next.
The other type of matrix is used in derivative stencils in 3D
RBF-FD calculations (Figure~\ref{fig:rbf3dnorcm} shows a $32$-point
stencil of a 3D $8^3$ grid).

@@

In the previous sections, we have explored the practical
implementation on an Intel Xeon Phi card of multiple derivative
operators acting on multiple vectors within the context of
RBF-FD. Each derivative has an associated sparse matrix with a fixed
number of nonzeros on each row. While computing a single derivative of
multiple functions is rather common, we accelerate the algorithm further 
by considering multiple
matrices, with identical sparsity patterns, acting on multiple vectors. 
We specialize our study to four matrices and four derivatives,
with 16 outputs, computed as a sum of outer products.

@@

In future work, we will examine the effect of larger stencil sizes and
double precision. The presented performance does not reach the
predicted peak and we believe that common techniques from SpMV such as
cache partitioning can be successfully applied. We will also integrate
the techniques presented in this paper within a fluid simulation using
RBF-FD. \ge{We will also look into developing a similar technique on the Kepler
family of GPUs, which offers swizzling and channeling operations.}
